\section{Evaluation\label{sec:results}}
To evaluate the usability of \theDeterminismChecker,
we applied it to the Randoop test
generator~\cite{PachecoLEB2007}.
Randoop is intended to be deterministic, when invoked on a deterministic
program~\cite{randoop-manual}.
%\footnote{Users of Randoop can pass in a different seed in order to
%    obtain a different deterministic output.\todo{Could cut this footnote
%      for space.}
%}
The developers had recently
spent two weeks of full-time work to
eliminate unintentional nondeterministic behavior, by repeatedly
running Randoop with verbose logging enabled and
looking for differences in logging output.
% (personal communication, 2019).

First, we wrote specifications for libraries Randoop uses, such as the JDK,
JUnit, and plume-util.
\todo{Say more about this.  In particular, the specification for output
  routines such as \<println> and \<printf> requires deterministic arguments.}
Second, we wrote type qualifiers in the Randoop source code to express its
determinism specification.\todo{Where did we get the determinism specification?}
Third, we ran
\theDeterminismChecker.  Each warning indicated a mismatch between the
specification and the implementation.  We addressed each warning by changing a
type qualifier, reporting a bug, or suppressing a false positive warning.
At the end of the process, Randoop's 24K lines of code contained 1348 annotations.
%98 \<@NonDet>,
%64 \<@OrderNonDet>,
%1083 \<@Det>,
%and 103 \<@PolyDet> annotations.
%\todo{The paper doesn't mention polymorphism.  Is the
%    breakdown of the different types of annotations interesting to readers?}
This is about 1 annotation per 17 lines of code,
% Justification for the claim about generics:
% There are about 1389 lines with uses of generics, according to the following:
% (rg "<.*>" "everything" (substitute-in-file-name "$test/randoop/src/main/java/"))
% (delete-matching-lines ": /?*\\(@Option\\|//\\|/?\\*\\|return \"\\|\"\\)" nil nil t)
% (replace-regexp " [<>] " "")
% (replace-regexp "[<>]=" "")
% (replace-regexp "[-=][<>]" "")
% (replace-sring "<init>" "")
% (delete-non-matching-lines "[<>]" nil nil t)
and is fewer than the uses
of Java generic types in Randoop.

\TheDeterminismChecker found \numRandoopBugs previously-unknown
nondeterminism bugs in Randoop.
The developers accepted our bug reports and committed fixes to the repository. 
Here are two examples
of bugs:

    \textbf{HashSet bug}: In Randoop's input (called the ``code under test''),
    if a type variable's lower or upper
    bound has a type parameter that the type variable itself does not have,
    then Randoop is nondeterministic.
    This situation does occur, even in Randoop's test suite.
    \Cref{fig:randoop-bug-hashset} shows the function that causes this nondeterministic behavior.
    \todo{The next part of this paragraph is boring.  It just restates
      exactly what is in the figure (but in a more verbose and less precise
      way).  There is no point in doing that.
      Instead, cut it or replace it by text that adds information.}
    We annotated the return type as \codeid{Det List<Det TypeVariable>} (line 160) since the method \codeid{getTypeParameters()}
    is expected to return deterministic results. The method creates a local \<HashSet>, \<parameters> 
    which we annotated as \codeid{OrderNonDet Set<Det TypeVariable>} (line 161). In the next 2 lines,
    the method adds an element to the set \<parameters> and returns an \<ArrayList> created with the
    contents of this set. Since the \<ArrayList> is created with an \<OrderNonDet> set as
    its argument, the type of \<ArrayList> is \<OrderNonDet>. The return type of \<getTypeParameters> 
    (\codeid{OrderNonDet List<Det TypeVariable>}) is not a subtype of the specified return type 
    (\codeid{Det List<Det TypeVariable>}) causing \theDeterminismChecker to issue a warning.
    The developers fixed this by changing a \<HashSet> to \<LinkedHashSet>
    (commit c975a9f7, shown\todo{This is not shown.} in \cref{fig:randoop-bug-hashset}).
    
    
%    \todo{Remove
%      the figure and instead give examples of specified code.  Or, show
%      just the ``before'' version of the code \emph{with annotations},
%      which will demonstrate how Randoop issues warnings.}.
%    \TheDeterminismChecker could detect this bug by assigning the type qualifier \<OrderNonDet>
%    to the newly created \<HashSet> (\<parameters>)
%    \todo{A
%      type-checker does not assign a type} 
%    \todo{This sentence is impossible to follow.  Each of
%      the two sentences
%      vaguely describes some difference between two versions of code.
%      Don't do that.  (A writing rule is ``show, don't tell''.)  Instead,
%      show the actual annotated code, and explain the warning that
%      \theDeterminismChecker issues.}
%    on line 161. The function returns an \<ArrayList> which gets the
%    determinism type \<Det>.
%    \todo{The following sentence is also unclear.  Who is the actor?  What
%      is the code?  Show it rather than giving this hard-to-follow explanation.}
%  Creating a \<Det ArrayList> with 
%    \<OrderNonDet> arguments violates the typing rules causing \theDeterminismChecker to report a bug.
    
    \textbf{Classpath bug}
    Randoop used the CLASSPATH environment variable in preference to the
    classpath passed on the command line.
    This can cause incorrect behavior
%    , both in Randoop's test suite and in the field,
    if a user sets the environment variable.
    The developers fixed the problem by changing Randoop to not read the environment variable
    (commit 330e3c56).




\section{Research Contributions\label{sec:contributions}}
\begin{enumerate}
  \item We designed a type system for expressing determinism properties.

  \item We implemented the analysis, as a pluggable type system for Java in a
    tool called \theDeterminismChecker.

  \item In a case study, we ran our analysis on a 24 KLOC project.
    \TheDeterminismChecker
    found 5 instances of nondeterminism that the developers had
    overlooked.
\end{enumerate}


%%  LocalWords:  util NonDet OrderNonDet Det PolyDet c975a9f7 e3c56 test''
% LocalWords:  println TypeVariable getTypeParameters
